Supernovae Time Profiles as a Portal to New Physics at Neutrino Telescopes

### 2024-01-23 paper/sections/conclusion.tex

\textbf{\textit{Results}}---
\begin{figure}[t!]
    \centering
    \includegraphics[width=0.47\textwidth]{figures/majoran_sensitivity}
    \caption{\textbf{\textit{Exclusion sensitivities for the Majoron case.}}
    The lines on this plot show the exclusion sensitivity for IceCube, and two next-generation neutrino experiments, Hyper-Kamiokande and DUNE.
    Additionally, we also show the reaches by SN cooling (blue) and by the lack of observation of high-energy neutrinos from SN1987A (grey) \cite{Fiorillo:2022cdq}, with the width of the lines quantifying uncertainties from SN modeling.
    }
    \label{fig:sensitivity}
\end{figure}
Looking for rapid raises within various time windows at neutrino telescope in the occurrence of future nearby SN explosion will provide exceptional tests of the parameter space. Assuming the SN event happens in the galaxy at a distance $D_{\rm SN}=\unit[10]{kpc}$, which is not unlikely~\cite{Reed:2005en,Rozwadowska:2020nab}, we consider the currently operating IceCube neutrino observatory as an example. 
The expected exclusion limit are shown in \cref{fig:sensitivity} (dark red) if no excess of hits on top of those from background noise and the standard neutrino flux were observed. JUNO expected to start taking data soon, while DUNE and Hyper-K requiring longer time to start operating, will be able to test the parameter of Majorons by looking for high energy neutrino following \cite{Fiorillo:2022cdq} if such SN event happens during their operation time. We present the estimated reaches of JUNO, DUNE and Hyper-K in \cref{fig:sensitivity} as xxx, red and orange lines, respectively, leaving aside careful detector analysis, together with current limits \cite{Fiorillo:2022cdq} from the aforementioned energy loss requirement (blue band), and non-observation of high energy neutrinos from SN 1987A (gray band) in \cref{fig:sensitivity}.

@@

Uncertainties on the analysis from different modeling SN explosion are minor, as studies in \cite{li2023old} and potential discrepancies between data and modeling are unresolved questions. Nevertheless we point out that the discrepancy could be within $2\sigma$ level which will only change our exclusion limit by a small factor.

\textbf{\textit{Outlook}}---
we can also study axion cases, mixing cases with this strategy. Pre-supernova bouncing.

>>>>

\textbf{\textit{Results}}---
\begin{figure}[t!]
    \centering
    \includegraphics[width=0.47\textwidth]{figures/majoran_sensitivity}
    \caption{\textbf{\textit{Exclusion sensitivities for the Majoron case.}}
    The lines on this plot show the exclusion sensitivity for IceCube, and two next-generation neutrino experiments, Hyper-Kamiokande and DUNE.
    Additionally, we also show the reaches by SN cooling (blue) and by the lack of observation of high-energy neutrinos from SN1987A (grey) \cite{Fiorillo:2022cdq}, with the width of the shading region quantifying uncertainties from SN modeling.
    }
    \label{fig:sensitivity}
\end{figure}
Looking for rapid raises within various time windows at neutrino telescope in the occurrence of future nearby SN explosion will provide exceptional tests of the parameter space. Assuming the SN event happens in the galaxy at a distance $D_{\rm SN}=\unit[10]{kpc}$, which is not unlikely~\cite{Reed:2005en,Rozwadowska:2020nab}, we consider the currently operating IceCube neutrino observatory as an example. 
The expected exclusion limit are shown in \cref{fig:sensitivity} (dark red) if no excess of hits on top of those from background noise and the standard neutrino flux were observed. DUNE and Hyper-K will be operating and be able to test the parameter of Majorons by looking for time-integrated high energy neutrinos following \cite{Fiorillo:2022cdq} if such SN event happens during their operation time. We present the estimated reaches of DUNE and Hyper-K in \cref{fig:sensitivity} as red and orange lines, respectively, leaving aside careful detector analysis, together with current limits \cite{Fiorillo:2022cdq} from the aforementioned energy loss requirement (blue band), and non-observation of high energy neutrinos from SN 1987A (gray band) in \cref{fig:sensitivity}.

@@

Uncertainties on the analysis from different modeling SN explosions are minor, as studies in \cite{li2023old} and potential discrepancies between data and modeling are unresolved questions.
Nevertheless, we point out that the discrepancy could be within $2\sigma$ level which will only change our exclusion limit by a small factor.
We also provide the code used in this work so that the impact of SN modeling uncertainties on the sensitivity of neutrino telescopes may be evaluated in the future.

\textbf{\textit{Outlook}}---
We demonstrated that neutrino telescope such as IceCube provides unprecedented probes of new physics, such as Majoron neutrinos and sterile neutrinos with magnetic portal, via measuring Supernova time profiles to a window as short as $\mathcal{O}(0.01)$ sec. Signatures outside the standard signal time window, including early-glow and after-glow, can be probed at neutrino telescope which enhances probes of new physics.
This analysis can also be extended to probing new particles such as sterile neutrinos with mixing portal where the resulting time profile requires more dedicated studies, especially for resonantly conversions.

### 2024-01-23 paper/sections/detector_response.tex

\textbf{\textit{Detector Response and Statistical Treatment}}---The IceCube Neutrino Observatory comprises 5,160 light-detecting digital optical modules (DOMs) buried in a cubic kilometer of the deep, transparent Antarctic ice sheet.
When neutrinos interact within or near the detector, the charged by-products emit photons that the DOMs can detect.
The DOMs are arranged on 86 strings of 60 DOMs with an inter-string distance between 70~m and 125~m.
This enables IceCube to resolve individual neutrinos with energies approximately ranging between a few GeV and a few PeV.
The neutrinos produced by SNe are far below this threshold, and the neutrinos cannot be individually resolved; however, the immense number of neutrinos produced in a SN produces an increase in the single-photon rate of the detector.
This increase in the rate of photons can be distinguished by the background caused by dark noise and radioactive activity in the DOM glass to enable the detection of galactic SNe events~\cite{Griswold:2023iwz}.

New neutrino physics will affect the development of a SN, and will distort the temporal structure of the photon signal seen in the IceCube detector.
Thus, one may look for excess events in certain time intervals as evidence of this new physics.
In this work, we simulate the light curve produced by a standard SN with a progenitor mass of $8.8~M_{\odot}$ and those produced by different BSM scenarios; see the left-hand panel of Fig.~\ref{fig:hits_and_likelihood} for example light curves.
We use the \texttt{ASTERIA}~\cite{spencer_griswold_2020_3926835} package in order to simulate the detector response to the different SNe scenarios as well as the thermal and radioactive noise in the DOMs. \yy{we need to provide a brief recasting of simulating the detector response, especially for different flavors.}
This package simulates light yields from coherent $\nu_{e}$ scattering off electrons in the ice, inverse beta decay of $\bar{\nu}_{e}$ on nuclei in the ice, and charged- and neutral-current interactions for $\nu_{\alpha}$.
In order to interface to this package, we use \texttt{SNEWPY}~\cite{baxter2021snewpy} package and, in particular, the \texttt{ParametrizedFlux} object.

>>>>

\textbf{\textit{Detector Response and Statistical Treatment}}---The IceCube Neutrino Observatory comprises 5,160 light-detecting digital optical modules (DOMs) buried in a cubic kilometer of the deep, transparent Antarctic ice sheet.
When neutrinos interact within or near the detector, the charged by-products emit photons that the DOMs can detect.
The DOMs are arranged on 86 strings of 60 DOMs with an inter-string distance between 70~m and 125~m.
This enables IceCube to resolve individual neutrinos with energies approximately ranging between a few GeV and a few PeV.
The neutrinos produced by SNe are far below this threshold, and the neutrinos cannot be individually resolved; however, the immense number of neutrinos produced in an SN produces an increase in the single-photon rate of the detector.
This increase in the rate of photons can be distinguished by the background caused by dark noise and radioactive activity in the DOM glass to enable the detection of galactic SNe events~\cite{Griswold:2023iwz}.

New neutrino physics will affect the development of a SN, and will distort the temporal structure of the photon signal seen in the IceCube detector.
Thus, one may look for excess events in certain time intervals as evidence of this new physics.
In this work, we simulate the light curve produced by a standard SN with a progenitor mass of $8.8~M_{\odot}$ and those produced by different BSM scenarios; see the left-hand panel of Fig.~\ref{fig:hits_and_likelihood} for example light curves.
We use the \texttt{ASTERIA}~\cite{spencer_griswold_2020_3926835} package in order to simulate the detector response to the different SNe scenarios as well as the thermal and radioactive noise in the DOMs. \yy{we need to provide a brief recasting of simulating the detector response, especially for different flavors.} \yy{contributions to bg}
This package simulates light yields from coherent $\nu_{e}$ scattering off electrons in the ice, inverse beta decay of $\bar{\nu}_{e}$ on nuclei in the ice, and charged- and neutral-current interactions for $\nu_{\alpha}$.
In order to interface to this package, we use \texttt{SNEWPY}~\cite{baxter2021snewpy} package and, in particular, the \texttt{ParametrizedFlux} object.

### 2024-01-23 paper/sections/introduction.tex

If, additionally, new physics particles, emitted from the SN core, can decay to neutrinos, even stronger limits can be set. In particular, for the standard case, neutrinos interact frequently in the core and eventually leave the star with $\mathcal{O}(10)$ MeV energy. New states after being produced can stream out without further interactions. Consequently, they typically exit with higher energies, subsequently decaying into neutrinos with an energy on the order of $\mathcal{O}(100)$ MeV. Limits can be set from the fact that such high energy neutrinos were not recorded from SN 1987A \cite{Kamiokande-II:1987idp,Bionta:1987qt,Baksan, Fiorillo:2022cdq, Brdar:2023tmi}. Once produced from SN, the new state can be non-relativistic and travels slowly, causing time delays to the arrival time of its daughter particle which was also recently employed \cite{Brdar:2023tmi} for a time window around one day. However, as being decoupled immediately, it is possible that new states can exit the supernova at an earlier time instead—an effect that has not been previously considered. These two effects in combination would cause advanced or delayed neutrino signals to the standard case.

>>>>

If, additionally, new physics particles, emitted from the SN core, can decay to neutrinos, even stronger limits can be set. In particular, for the standard case, neutrinos interact frequently in the core and eventually leave the star with $\mathcal{O}(10)$ MeV energy. New states after being produced can stream out without further interactions. Consequently, they typically exit with higher energies, subsequently decaying into neutrinos with an energy on the order of $\mathcal{O}(100)$ MeV. Limits can be set from the fact that such high energy neutrinos were not recorded from SN 1987A \cite{Kamiokande-II:1987idp,Bionta:1987qt,Baksan, Fiorillo:2022cdq, Brdar:2023tmi}. Once produced from SN, the new state can be non-relativistic and travels slowly, causing time delays to the arrival time of its daughter particle which was also recently employed \cite{Brdar:2023tmi} for a time window around one day. However, as being decoupled immediately, it is possible that new states can exit the supernova at an earlier time instead—an effect that has not been previously considered. This early-glow signal can also be induced by light new states produced from $\nu_e$ scattering inside the supernova given that $\nu_e$ burst happens at an earlier time than the other neutrino flavors \cite{Brdar_2018}.

### 2024-01-23 paper/sections/model.tex

\begin{figure}[t!]
    \centering
    \includegraphics[width=0.47\textwidth]{figures/majoran_fluxes.pdf}
    \caption{\textit{\textbf{Neutrino flux from Standard Model production and two majoron hypotheses}.
    We choose the parameters $g_\phi = 10^{-10.2}$ and $g_\phi = 10^{-11.7}$ for $m_\phi = \unit[1]{MeV}$ and $m_\phi = \unit[100]{MeV}$, respectively.}
    }
    \label{fig:fluxes}
\end{figure}

For $D_{\rm SN} = \unit[10]{kpc}$ and $\delta t = \unit[100]{s}$ while larger values of $\delta t$ do not affect the limits obtained, we show in \cref{fig:fluxes} the differential fluxes of daughter neutrinos at different time for two benchmarks of the Majoron case, in comparison to the standard neutrino flux.
We can observe that the energy of neutrinos from Majoron decay extends to $\unit[100]{MeV}$ and above, as expected. For heavy Majoron case, the resulting neutrino flux typically is larger at later time, manifesting the time delay from slowly moving Majorons. 
The flux for light Majoron case contain very rich physics. Majorons, generating neutrinos with $E_\nu \geq \unit[150]{MeV}$, are nearly-relativistic with its daughter particles moving in parallel. In such a case, for couplings comparable to or larger than the one shown in \cref{fig:fluxes}, $\Delta t \sim L_d (1/\beta -1)\lesssim \unit[0.01]{sec}$, with $L_d = E_\phi/m_\phi \Gamma^{-1}_\phi \beta$ being the decay length of $\phi$ for the given $\phi$ energy $E_\phi$. The time dependence of neutrino flux thus are mainly inherited from the Majorons upon production, which exhibits in \cref{fig:fluxes} larger fluxes from producing Majorons from the denser and hotter core of SN at earlier time.
For neutrinos with lower energy, i.e $E_\nu \lesssim \unit[40]{MeV}$, the mother Majorons can be non-relativistic, resulting in $\Delta t \gtrsim \unit[0.04]{sec}$. This non-negligible time delay would shift the neutrino flux to later time, lowering the fluxes at earlier time as shown in \cref{fig:fluxes}. For the heavy mass benchmark in \cref{fig:fluxes}, we have $\Delta t \gtrsim 10 sec$, which would yield the active neutrino flux at sizeable $t\sim \mathcal{sec}$.

@@

\begin{figure*}
    \centering
    \includegraphics[width=0.95\textwidth]{figures/hits_and_likelihood.pdf}
    \caption{\textbf{\textit{Test statistic as a function of $t_{\mathrm{start}}$ and $\Delta$t.}
    }}
    \label{fig:hits_and_likelihood}
\end{figure*}

>>>>

\begin{figure}[t!]
    \centering
    \includegraphics[width=0.47\textwidth]{figures/majoran_fluxes.pdf}
    \caption{\textit{\textbf{Flux of $\bar{\nu}_{e}$ from Standard Model production and two majoron hypotheses}}.
    We choose the parameters $g_\phi = 10^{-10.2}$ and $g_\phi = 10^{-11.8}$ for $m_\phi = \unit[1]{MeV}$ and $m_\phi = \unit[200]{MeV}$, respectively.
    }
    \label{fig:fluxes}
\end{figure}

For $D_{\rm SN} = \unit[10]{kpc}$ and $\delta t = \unit[100]{s}$ while larger values of $\delta t$ do not affect the limits obtained, we show in \cref{fig:fluxes} the differential fluxes of daughter neutrinos at different time for two benchmarks of the Majoron case, in comparison to the standard neutrino flux.
We can observe that the energy of neutrinos from Majoron decay extends to $\unit[100]{MeV}$ and above, as expected. For heavy Majoron case, the resulting neutrino flux typically is larger at later time, manifesting the time delay from slowly moving Majorons. 
The flux for light Majoron case contain very rich physics. Majorons, generating neutrinos with $E_\nu \geq \unit[150]{MeV}$, are nearly-relativistic with its daughter particles moving in parallel. In such a case, for couplings comparable to or larger than the one shown in \cref{fig:fluxes}, $\Delta t \sim L_d (1/\beta -1)\lesssim \unit[0.01]{sec}$, with $L_d = E_\phi/m_\phi \Gamma^{-1}_\phi \beta$ being the decay length of $\phi$ for the given $\phi$ energy $E_\phi$. The time dependence of neutrino flux thus are mainly inherited from the Majorons upon production, which exhibits in \cref{fig:fluxes} larger fluxes from producing Majorons from the denser and hotter core of SN at earlier time.
For neutrinos with lower energy, i.e $E_\nu \lesssim \unit[40]{MeV}$, the mother Majorons can be non-relativistic, resulting in $\Delta t \gtrsim \unit[0.04]{sec}$. This non-negligible time delay would shift the neutrino flux to later time, lowering the fluxes at earlier time as shown in \cref{fig:fluxes}. For the heavy mass benchmark in \cref{fig:fluxes}, we have $\Delta t \gtrsim \unit[10]{sec}$, which would yield the active neutrino flux at sizeable $t$.

@@

\begin{figure*}
    \centering
    \includegraphics[width=0.95\textwidth]{figures/hits_and_likelihood.pdf}
    \caption{\textbf{\textit{Timing profile of hits and test statistic as a function of the timing window.}}
    These are the same points in the parameter space as Fig.~\ref{fig:fluxes}.
    }
    \label{fig:hits_and_likelihood}
\end{figure*}
